\textsc{\wobble: a data-driven \changed{analysis technique for time-series stellar spectra}}

### 2019-08-19 paper/paper.tex
slightly better math notation

We take the data to be the $M \times N$ matrix $y$, where each entry $y_{m,n}$ is the logarithm of the observed flux for pixel $m$ of $M$ at epoch $n$ of $N$. 
We also have a corresponding $M \times N$ matrix of wavelength solutions \changed{which we call $\xi$, where each entry $\xi_{m,n}$ is the logarithm of the wavelength for pixel $m$ at epoch $n$}.

\changed{For each data column $y_n$, our model prediction $f_n$ can be treated as the sum of stellar and telluric contributions:
\begin{equation}
f_n = f_{\star,n} + f_{t,n} + \textrm{noise}.
\end{equation}}

\changed{The stellar spectrum contribution is: 
\begin{equation}
f_{\star,n} = P(\xi_n, \xi_{\star}, v_{obs,n}) \cdot \mu_{\star},
\label{eqn:star}
\end{equation}
where $\mu_{\star}$ is a spectral template of log-fluxes and 
$\xi_{\star}$ is the corresponding vector of template log-wavelengths. 
The exact values of $\xi_{\star}$ can be chosen somewhat arbitrarily under the conditions that the grid has uniform spacing $\Delta\xi_{\star}$, covers the entire wavelength range of the data, and is over-sampled with regards to the observed spectrum's wavelength grid.
$P$ is a linear operator whose function is to apply a Doppler shift by observed velocity $v_{obs,n}$ and
interpolate $\mu_{\star}$ from the $\xi_{\star}$ template grid to the $\xi_n$ data grid.
Each entry of the $P$ matrix can be defined by a sum of weighted indicator functions as follows:
\begin{equation}
\begin{split}
 P_{i,j} =  &\ \Big(\frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}}\Big) \mathbb{1}\big(0 < \frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}} < 1\big) \\
  &+ \Big(1 - \frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}}\Big) \mathbb{1}\big({-1} < \frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}} < 0\big), \\
 \end{split}
\end{equation}
where $\xi'_{\star}$ is the Doppler-shifted template grid:
\begin{equation}
 \xi'_{\star}(v) = \xi_{\star} + D(v) \equiv \xi_{\star} + \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
\end{equation}}
Because $P$ is quite sparse, in practice we do not instantiate the full matrix when performing calculations. 
\changed{At this point, it is also useful to note that $P$ could, in principle, encode an instrumental line spread function (\LSF). 
We return to this point in Section \ref{s:future}.}

@@

The telluric spectrum contribution is:
\changed{\begin{equation}
f_{t,n} =  A_n(\mu_{t} + W_{t} \cdot z_n).
\end{equation}}
In addition to its mean spectrum $\mu_t$, the telluric component also includes a time-dependent term assembled from two variables: $W_t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$. 
\changed{$W_t$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $M'$ is the length of some vector of template wavelengths $\xi_{t}$ and $K$ is the number of basis vectors used.\footnote{While we refer to these as basis vectors for simplicity, note that they are not actually constrained to be orthogonal.}} 
For the purposes of this work we found good performance with $K$ set to 3, but this may vary for other applications. 
\changed{The template wavelength grid $\xi_{t}$ should have similar properties to $\xi_{\star}$.} 
Finally, the net telluric spectrum (mean + time-variable components) is weighted by the airmass at the time of observation $A_n$, a known quantity.

We evaluate a log-likelihood for each epoch
\changed{
\begin{equation}
\ln \mathcal{L}_n = -\frac{1}{2} (y_{n} - f_{n})^T C_{n}^{-1} (y_{n}-f_{n}),
\label{eqn:lnlike}
\end{equation}}
comparing our model \changed{$f$} to the data $y$, with $C$ representing the covariance matrix of uncertainties for the data.

@@

L1 normalization adds a term to the log-likelihood that takes the form:
\begin{equation}
\Lone{\lambda}{p} \equiv -\lambda \sum_{i} | p_{i} | ,
\end{equation}
where $p$ is the vector of parameters to be normalized (in this case $\mu_{\star}$, $\mu_{t}$, or $W_{t}$) and $\lambda$ is the regularization amplitude.

@@

The effectiveness of the regularization depends sensitively on the value of $\lambda$ used: if $\lambda$ is too high, real features will be lost as the parameters are forced to zero, whereas setting $\lambda$ too low will make the regularization ineffective, leaving the model vulnerable to overfitting.

We set regularization amplitudes for \wobble using a cross-validation scheme. 
In brief, we randomly select $10-15$\% of the total epochs to set aside as a validation set and, using some value of $\lambda$, run the model optimization on the remaining epochs (the training set). 
The resulting best-fit spectral templates and basis vectors are taken as fixed and the time-dependent terms only (\RVs and basis weights) are optimized for the validation epochs. 
The $\chi^2$ for the validation epochs can then be adopted as a goodness-of-fit measurement for the $\lambda$ value, and the procedure is repeated for different $\lambda$s to choose the best regularization amplitude. 
In theory, we might wish to regularize using an optimization metric based on \RV accuracy rather than $\chi^2$ over pixels; however, the lack of known ``ground truth'' in the \RV behavior of stars makes this not currently feasible.

@@

With regularization included, the final model likelihood to be optimized is:
\begin{equation}
\begin{split}
\ln \mathcal{L} = & -\frac{1}{2} \sum_{n} (y_{n} - f_{n})^T C_{n}^{-1} (y_{n}-f_{n})  \\
 & + \Lone{\lambda_1}{\mu_{\star}} + \Ltwo{\lambda_2}{\mu_{\star}}   + \Lone{\lambda_3}{\mu_{t}} + \Ltwo{\lambda_4}{\mu_{t}} \\
 &  + \Lone{\lambda_5}{W_t}  + \Ltwo{\lambda_6}{W_t} + \Lone{1.0}{z} ,
\end{split}
\end{equation}
%\todo{BEDELL: THIS IS WRONG. We were supposed to set $\lambda_5$ to unity. My (Hogg's) bad. However, I wrote the following to justify our WRONG choice!}
where the regularization amplitude on the basis weights is arbitrarily set to unity, and all other regularization amplitudes $\lambda$ are set by grid searches using the above-described validation procedure.
The basis-weight regularization amplitude can be set arbitrarily to unity because there is a perfect degeneracy between amplifying the basis vectors and attenuating the basis weights.
This regularization choice breaks that degeneracy.
The basis weights are regularized to encourage our desired outcome where the mean telluric spectrum contains as much telluric information as possible, while the variable basis picks up only the necessary time-variable changes.

@@

As an initial guess, we set the star to be stationary, e.g. $v_{obs} = $ \BERV at all epochs. 
We may then initialize the stellar spectrum $\mu_{\star}$ by Doppler-shifting the data and calculating the median flux across \BERV-corrected spectra in bins at each model wavelength \changed{$\xi_{\star}$}. 
The telluric template is initialized similarly by using the residuals after the stellar contribution has been removed, again binning in model wavelength (this time without applying any Doppler shift to the spectra) and taking the median values of each bin. 
Finally, we initialize the telluric spectrum's basis vectors \changed{$W_t$} and weights \changed{$z_t$} by performing Principal Component Analysis (\acronym{PCA}) on the residuals after both the stellar spectrum and the mean telluric spectrum have been removed. 
The $K$ highest eigenweights and their corresponding eigenvectors are taken as \changed{the starting guess for} the basis weights and vectors.

Once all parameters are initialized, we optimize them iteratively. 
The likelihood function is maximized first by varying the stellar and telluric templates (including the telluric basis vectors). 
This step is a convex optimization, meaning that a global optimum for the template parameters should be reached under the condition of fixed time-dependent variables. 
Next, the templates are held fixed and the time-dependent parameters (stellar \RVs and telluric basis weights) are varied to maximize the likelihood function again. 
Technically the velocities are location parameters and their optimization is not convex.
However, once the velocities are known to a small fraction of a pixel (which they usually are in practice, given an accurate estimation of the \BERV), the linearized problem becomes convex at each iteration.
We repeat this procedure, optimizing spectra and time-dependent parameters in turn, until the likelihood appears to converge, typically within 100 iterations.
This iterative procedure is equivalent to a full simultaneous optimization in the limit of many iterations to convergence.

After convergence, we estimate the uncertainties on the parameters by
approximating the likelihood function near its maximum as Gaussian where the
covariance matrix is the negative inverse Hessian (or second derivative
matrix) of the log likelihood function with respect to the parameters.

@@

In the above analyses, telluric spectra were inferred independently for the 51 Peg and \Mdwarf data sets. 
However, in principle the telluric features should be common to all \HARPS spectra, an assumption that we used for the case of HD 189733.

We show a comparison between 51 Peg and \Mdwarf telluric fits in Figure \ref{fig:telluric_comparison}. 
Although they were fit using entirely different data sets with dissimilar stellar spectra, the resulting telluric fits are indeed extremely similar. 
This is generally true for all wavelength regions in which tellurics are detected. 
The only regime in which the comparison fails is for low-\SNR spectral orders of the \Mdwarf data, where the telluric spectra are featureless due to the strong regularization, as discussed in Section \ref{s:Mdwarf}.

@@

Another way in which the \wobble model may fail for some regimes is in its treatment of spectral variability.

We have assumed that the telluric spectrum is variable in a low-dimensional sense only. 
While this assumption has a basis in the physical understanding of telluric lines, which are produced by only a small number of molecular species, it is not guaranteed that this reasoning translates into low variability in the current parameterization. 
We may instead wish to change the spectral representation so that additional physics is included, as discussed further in Section \ref{s:improvements}. 
Alternatively, we could keep the current representation but fit the telluric model to the residuals away from some physics-based model.

@@

This work has made use of data from the European Space Agency (ESA) mission
{\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia}
Data Processing and Analysis Consortium (DPAC,
\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
has been provided by national institutions, in particular the institutions
participating in the {\it Gaia} Multilateral Agreement.}

>>>>

We take the data to be the $M \times N$ matrix \mmatrix{y}, where each entry $y_{m,n}$ is the logarithm of the observed flux for pixel $m$ of $M$ at epoch $n$ of $N$. 
We also have a corresponding $M \times N$ matrix of wavelength solutions \changed{which we call \mmatrix{\xi}, where each entry $\xi_{m,n}$ is the logarithm of the wavelength for pixel $m$ at epoch $n$}.

\changed{For each data column \mvec{y_n}, our model prediction \mvec{f_n} can be treated as the sum of stellar and telluric contributions:
\begin{equation}
\mvec{f_n} = \mvec{f_{\star,n}} + \mvec{f_{t,n}} + \textrm{noise}.
\end{equation}}

\changed{The stellar spectrum contribution is: 
\begin{equation}
\mvec{f_{\star,n}} = \mmatrix{P}(\mvec{\xi_n}, \mvec{\xi_{\star}}, v_{obs,n}) \cdot \mvec{\mu_{\star}},
\label{eqn:star}
\end{equation}
where \mvec{\mu_{\star}} is a spectral template of log-fluxes and 
\mvec{\xi_{\star}} is the corresponding vector of template log-wavelengths. 
The exact values of \mvec{\xi_{\star}} can be chosen somewhat arbitrarily under the conditions that the grid has uniform spacing $\Delta\xi_{\star}$, covers the entire wavelength range of the data, and is over-sampled with regards to the observed spectrum's wavelength grid.
\mmatrix{P} is a linear operator whose function is to apply a Doppler shift by observed velocity $v_{obs,n}$ and
interpolate \mvec{\mu_{\star}} from the \mvec{\xi_{\star}} template grid to the \mvec{\xi_n} data grid.
Each entry of the \mmatrix{P} matrix can be defined by a sum of weighted indicator functions (where an indicator function is denoted here as $\mathbb{1}(x)$ and is defined to have value 1 when condition $x$ is fulfilled and zero otherwise):
\begin{equation}
\begin{split}
 P_{i,j} =  &\ \Big(\frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}}\Big) \mathbb{1}\big(0 < \frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}} < 1\big) \\
  &+ \Big(1 - \frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}}\Big) \mathbb{1}\big({-1} < \frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}} < 0\big), \\
 \end{split}
\end{equation}
where \mvec{\xi'_{\star}} is the Doppler-shifted template grid:
\begin{equation}
 \mvec{\xi'_{\star}}(v) = \mvec{\xi_{\star}} + D(v) \equiv \mvec{\xi_{\star}} + \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
\end{equation}}
Because \mmatrix{P} is quite sparse, in practice we do not instantiate the full matrix when performing calculations. 
\changed{At this point, it is also useful to note that \mmatrix{P} could, in principle, encode an instrumental line spread function (\LSF). 
We return to this point in Section \ref{s:future}.}

@@

The telluric spectrum contribution is:
\changed{\begin{equation}
\mvec{f_{t,n}} =  A_n(\mvec{\mu_{t}} + \mmatrix{W_{t}} \cdot \mvec{z_n}).
\end{equation}}
In addition to its mean spectrum \mvec{\mu_t}, the telluric component also includes a time-dependent term assembled from two variables: \mmatrix{W_t}, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by \mvec{z_n} to form the spectral contribution at epoch $n$. 
\changed{\mmatrix{W_t} has the shape $M' \times K$ and \mvec{z_n} is a $K$-vector, where $M'$ is the length of some vector of template wavelengths \mvec{\xi_{t}} and $K$ is the number of basis vectors used.\footnote{While we refer to these as basis vectors for simplicity, note that they are not actually constrained to be orthogonal.}} 
For the purposes of this work we found good performance with $K$ set to 3, but this may vary for other applications. 
\changed{The template wavelength grid \mvec{\xi_{t}} should have similar properties to \mvec{\xi_{\star}}.} 
Finally, the net telluric spectrum (mean + time-variable components) is weighted by the airmass at the time of observation $A_n$, a known quantity.

We evaluate a log-likelihood for each epoch
\changed{
\begin{equation}
\ln \mathcal{L}_n = -\frac{1}{2} (\mvec{y_{n}} - \mvec{f_{n}})^T \mvec{C_{n}}^{-1} (\mvec{y_{n}} - \mvec{f_{n}}),
\label{eqn:lnlike}
\end{equation}}
comparing our model \changed{\mmatrix{f}} to the data \mmatrix{y}, with \mmatrix{C} representing the covariance matrix of uncertainties for the data.

@@

L1 normalization adds a term to the log-likelihood that takes the form:
\begin{equation}
\Lone{\lambda}{p} \equiv -\lambda \sum_{i} | p_{i} | ,
\end{equation}
where \mvec{p} is the vector of parameters to be normalized (in this case \mvec{\mu_{\star}}, \mvec{\mu_{t}}, or \mmatrix{W_{t}}) and $\lambda$ is the regularization amplitude.

@@

The effectiveness of the regularization depends sensitively on the value of $\lambda$ used: if $\lambda$ is too high, real features will be lost as the parameters are forced to zero, whereas setting $\lambda$ too low will make the regularization ineffective, leaving the model vulnerable to overfitting. 
We set regularization amplitudes for \wobble using a cross-validation scheme. 
In brief, we randomly select $10-15$\% of the total epochs to set aside as a validation set and, using some value of $\lambda$, run the model optimization on the remaining epochs (the training set). 
The resulting best-fit spectral templates and basis vectors are taken as fixed and the time-dependent terms only (\RVs and basis weights) are optimized for the validation epochs. 
The $\chi^2$ for the validation epochs can then be adopted as a goodness-of-fit measurement for the $\lambda$ value, and the procedure is repeated for different $\lambda$s to choose the best regularization amplitude. 
In theory, we might wish to regularize using an optimization metric based on \RV accuracy rather than $\chi^2$ over pixels; however, the lack of known ``ground truth'' in the \RV behavior of stars makes this not currently feasible.

@@

With regularization included, the final model likelihood to be optimized is:
\begin{equation}
\begin{split}
\ln \mathcal{L} = & -\frac{1}{2} \sum_{n} (\mvec{y_{n}} - \mvec{f_{n}})^T \mvec{C_{n}}^{-1} (\mvec{y_{n}} - \mvec{f_{n}})  \\
 & + \Lone{\lambda_1}{\mu_{\star}} + \Ltwo{\lambda_2}{\mu_{\star}}   + \Lone{\lambda_3}{\mu_{t}} + \Ltwo{\lambda_4}{\mu_{t}} \\
 &  + \Lone{\lambda_5}{W_t}  + \Ltwo{\lambda_6}{W_t} + \Lone{1.0}{z} ,
\end{split}
\end{equation}
%\todo{BEDELL: THIS IS WRONG. We were supposed to set $\lambda_5$ to unity. My (Hogg's) bad. However, I wrote the following to justify our WRONG choice!}
where the regularization amplitude on the basis weights is arbitrarily set to unity, and all other regularization amplitudes $\lambda$ are set by grid searches using the above-described validation procedure.
The basis-weight regularization amplitude can be set arbitrarily to unity because there is a perfect degeneracy between amplifying the basis vectors and attenuating the basis weights.
This regularization choice breaks that degeneracy.
The basis weights are regularized to encourage our desired outcome where the mean telluric spectrum contains as much telluric information as possible, while the variable basis picks up only the necessary time-variable changes.

@@

As an initial guess, we set the star to be stationary, e.g. $v_{obs} = $ \BERV at all epochs. 
We may then initialize the stellar spectrum \mvec{\mu_{\star}} by Doppler-shifting the data and calculating the median flux across \BERV-corrected spectra in bins at each model wavelength \changed{\mvec{\xi_{\star}}}. 
The telluric template is initialized similarly by using the residuals after the stellar contribution has been removed, again binning in model wavelength (this time without applying any Doppler shift to the spectra) and taking the median values of each bin. 
Finally, we initialize the telluric spectrum's basis vectors \changed{\mmatrix{W_t}} and weights \changed{\mvec{z_t}} by performing Principal Component Analysis (\acronym{PCA}) on the residuals after both the stellar spectrum and the mean telluric spectrum have been removed. 
The $K$ highest eigenweights and their corresponding eigenvectors are taken as \changed{the starting guess for} the basis weights and vectors.

Once all parameters are initialized, we optimize them iteratively. 
The likelihood function is maximized first by varying the stellar and telluric templates (including the telluric basis vectors). 
This step is a convex optimization, meaning that a global optimum for the template parameters should be reached under the condition of fixed time-dependent variables. 
Next, the templates are held fixed and the time-dependent parameters (stellar \RVs and telluric basis weights) are varied to maximize the likelihood function again. 
Technically the velocities are location parameters and their optimization is not convex.
However, once the velocities are known to a small fraction of a pixel (which they usually are in practice, given an accurate estimation of the \BERV), the linearized problem becomes convex at each iteration.
We repeat this procedure, optimizing spectra and time-dependent parameters in turn, until the likelihood appears to converge, typically within 100 iterations.
This iterative procedure is equivalent to a full simultaneous optimization in the limit of many iterations to convergence.
After convergence, we estimate the uncertainties on the parameters by
approximating the likelihood function near its maximum as Gaussian where the
covariance matrix is the negative inverse Hessian (or second derivative
matrix) of the log likelihood function with respect to the parameters.

@@

In the above analyses, telluric spectra were inferred independently for the 51 Peg and \Mdwarf data sets. 
However, in principle the telluric features should be common to all \HARPS spectra, an assumption that we used for the case of HD 189733. 
\changed{As a test of this assumption, we made a comparison between the telluric template spectra derived from the 51 Peg and \Mdwarf data (Figure \ref{fig:telluric_comparison}).} 
Although they were fit using entirely different data sets with dissimilar stellar spectra, the resulting telluric fits are indeed extremely similar. 
This is generally true for all wavelength regions in which tellurics are detected. 
The only regime in which the comparison fails is for low-\SNR spectral orders of the \Mdwarf data, where the telluric spectra are featureless due to the strong regularization, as discussed in Section \ref{s:Mdwarf}.

@@

Another way in which the \wobble model may fail for some regimes is in its treatment of spectral variability.
We have assumed that the telluric spectrum is variable in a low-dimensional sense only. 
While this assumption has a basis in the physical understanding of telluric lines, which are produced by only a small number of molecular species, it is not guaranteed that this reasoning translates into low variability in the current parameterization. 
We may instead wish to change the spectral representation so that additional physics is included, as discussed further in Section \ref{s:improvements}. 
Alternatively, we could keep the current representation but fit the telluric model to the residuals away from some physics-based model.

@@

This work has made use of data from the European Space Agency (ESA) mission
{\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia}
Data Processing and Analysis Consortium (DPAC,
\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
has been provided by national institutions, in particular the institutions
participating in the {\it Gaia} Multilateral Agreement.

Work by B.T.M. was performed under contract with the Jet
Propulsion Laboratory (JPL) funded by NASA through
the Sagan Fellowship Program executed by the NASA
Exoplanet Science Institute.}
